Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-07-26 submission/manuscript.Rmd
edit methods

## Classifying tissue-enriched and germline-enriched genes
Tissue-enriched differentially expressd genes were determined by their classification in a previously published list of genes enriched in 17 male and female mouse tissues[@liComprehensiveMouseTranscriptomic2017]. This study defined expression as greather than 1 Fragments Per Kilobase of transcript per Million mapped read (FPKM) and tissue enrichment as at least 4-fold higher expression than any other tissue.

@@

Emrbyonic stem cells (ESCs) were initially cultured in primed ESC (pESC) media consisting of KnockOut DMEM (Gibco#10829–018), fetal bovine serum (Gibco#A5209501), KnockOut serum replacement (Invitrogen#10828–028), Glutamax (Gibco#35050-061), Anti-Anti (Gibco#15240-062), MEM Non-essential amino acids (Gibco#11140-050), and beta-mercaptoethanol (Sigma#M7522). They were then transitioned into ground-state "naive" ESCs (nESCs) by culturing for four passages in N2B27 media containing DMEM/F12 (Gibco#11330–032), Neurobasal media (Gibco#21103–049), Gluamax, Anti-Anti, N2 supplement (Invitrogen#17502048), and B27 supplement without vitamin A (Invitrogen#12587-010), and beta-mercaptoethanol. Both pESC and nESC media were supplemented with the GSK3 inhibitor CHIR99021 (Sigma #SML1046-5MG), the MEK inhibitor PD0325901 (Sigma #PZ0162-5MG), and leukemia inhibitory factor (LIF, Millipore#ESG1107).

@@

## Immunocytochemistry (ICC)
ICC of DAZL in EpiLCs was performed by washing cells thrice with phosphobuffered saline (PBS), fixing cells in 4% paraformaldehyde, washing thrice in PBS, blocking in PBS containing 0.3% Triton X-100, and 5% fetal bovine serum for 1 hour, washing thrice with PBS, and incubating in primary antibody (Rabbit anti DAZL, abcam ab34139, 1:200) in the blocking solution overnight at 4C. The next day cells were rinsed thrice with PBS, incubated in secondary antibody (Alexafluor 488 Invitrogen #710369, 1:1000) in blocking buffer, rinsed thrice in PBS, and then imaged. Images were taken blinded for genotype, chosen based on similar levels of DAPI signal, and then quantified via ImageJ.

## RNA sequencing 
### Published datasets
All published datasets are available at the Gene Expression Omnibus (GEO) https://www.ncbi.nlm.nih.gov/geo. Previously published RNA sequencing datasets analyzed in this study included the male wild-type and _Kdm5c_-KO adult amygdala and hippocampus[@vallianatosMutuallySuppressiveRoles2020] (available at GEO: GSE127722) and male wild-type and _Kdm5c_-KO EpiLCs[@samantaActivationXistEvolutionarily2022] (available at GSE: GSE96797).

### Alignment and analysis
After ensuring read quality via FastQC (v0.11.8), reads were then mapped to the mm10 _Mus musculus_ genome (Gencode) using STAR (v2.5.3a), during which we removed duplicates and kept only uniquely mapped reads. Count files were generated by FeatureCounts (Subread v1.5.0), and BAM files were converted to bigwigs using deeptools (v3.1.3) and visualized by the UCSC genome browser. RStudio (v3.6.0) was then used to analyze counts files by DESeq2 (v1.26.0)[@loveModeratedEstimationFold2014a] to identify differentially expressed genes (DEGs) with a q-value (p-adjusted via FDR/Benjamini–Hochberg correction) less than 0.1 and a log2 fold change greater than 0.5. For all DESeq2 analyses, log2 fold changes were calculated with lfcShrink using the ashr package[@stephensFalseDiscoveryRates2016]. MA-plots were generated by ggpubr (v0.4.0), and Eulerr diagrams were generated by eulerr (v6.1.1). Boxplots and scatterplots were generated by ggpubr (v0.6.0) and ggplot2 (v3.3.2). Heatmaps of gene expression were generated using the base R functions scale and hclust and visualized using the R package ComplexHeatmap (v2.12.1). The Upset plot was generated via the package UpSetR (v1.4.0)[@conwayUpSetRPackageVisualization2017]. Gene ontology (GO) analyses were performed by the R package enrichPlot (v1.16.2) using the biological processes setting.

## Chromatin immunoprecipitation followed by DNA sequecing (ChIP-seq)
We analyzed our previously published KDM5C ChIP-seq experiments in primary neuron cultures from the cortex and hippocampus[@iwaseMouseModelXlinked2016] (available at GEO: GSE61036) and EpiLCs[@samantaActivationXistEvolutionarily2022] (available at GEO: GSE96797). ChIP-seq of histone 3 lysine 4 dimethylation in male EpiLCs[@samantaActivationXistEvolutionarily2022] is also available at GEO: GSE96797. ChIP-seq of histone 3 lysine 4 trimethylation in the male amygdala[@vallianatosMutuallySuppressiveRoles2020] are avilable at GEO: GSE127817.

ChIP-seq reads were aligned to mm10 using Bowtie1 (v1.1.2) allowing up to two mismatches. Only uniquely mapped reads were used for analysis. Peaks were called using MACS2 software (v2.1.0.20140616) using input BAM files for normalization, with filters for a q-value < 0.1 and a fold enrichment > 1. Common peak sets were obtained in R via DiffBind (v3.6.5), and count tables for the common peaks were generated with the Bedtools multicov command. We removed “black-listed” genomic regions that often give aberrant signals. Peak proximity to genome annotations was determined by ChIPSeeker (v1.32.1). Enriched motifs were identified using HOMER[@heinzSimpleCombinationsLineageDetermining2010]. Gene ontology (GO) analyses were performed by the R package enrichPlot (v1.16.2) using the biological processes setting. Average binding across the genome was visualized using deeptools (v3.1.3). Bigwigs were visualized using the UCSC genome browser.

@@

## Data availability

@@

## Discussion notes

>>>>

## Classifying tissue-enriched and germline-enriched genes
Tissue-enriched differentially expressd genes (DEGs) were determined by their classification in a previously published dataset from 17 male and female mouse tissues[@liComprehensiveMouseTranscriptomic2017]. This study defined tissue expression as greather than 1 Fragments Per Kilobase of transcript per Million mapped read (FPKM) and tissue enrichment as at least 4-fold higher expression than any other tissue.

@@

Emrbyonic stem cells (ESCs) were initially cultured in primed ESC (pESC) media consisting of KnockOut DMEM (Gibco#10829–018), fetal bovine serum (Gibco#A5209501), KnockOut serum replacement (Invitrogen#10828–028), Glutamax (Gibco#35050-061), Anti-Anti (Gibco#15240-062), MEM Non-essential amino acids (Gibco#11140-050), and beta-mercaptoethanol (Sigma#M7522). They were then transitioned into ground-state "na&#239;ve" ESCs (nESCs) by culturing for four passages in N2B27 media containing DMEM/F12 (Gibco#11330–032), Neurobasal media (Gibco#21103–049), Gluamax, Anti-Anti, N2 supplement (Invitrogen#17502048), and B27 supplement without vitamin A (Invitrogen#12587-010), and beta-mercaptoethanol. Both pESC and nESC media were supplemented with the GSK3 inhibitor CHIR99021 (Sigma #SML1046-5MG), the MEK inhibitor PD0325901 (Sigma #PZ0162-5MG), and leukemia inhibitory factor (LIF, Millipore#ESG1107).

@@

## Immunocytochemistry (ICC)
ICC of DAZL in EpiLCs was performed by first growing cells on coverslips. Cells were then washed thrice with phosphobuffered saline (PBS), fixed in 4% paraformaldehyde, washed thrice in PBS, and blacked in PBS containing 0.3% Triton X-100, and 5% fetal bovine serum for 1 hour. They were then washed thrice with PBS and incubating in primary antibody (Rabbit anti DAZL, abcam ab34139, 1:200) in the blocking solution overnight at 4°C with gentle rocking. The next day, cells were rinsed thrice with PBS, and incubated in secondary antibody (Alexafluor 488 Invitrogen #710369, 1:1,000) with DAPI (1:1,000) in blocking buffer for 1 hour at room temperature. Coverslips were then rinsed thrice in PBS, and mounted onto slides using Prolong Gold (Invitrogen #P36930). Images were taken blinded for genotype, chosen based on similar levels of DAPI signal, and quantified via ImageJ before unblinding.

## RNA sequencing (RNA-seq)
After ensuring read quality via FastQC (v0.11.8), reads were then mapped to the mm10 _Mus musculus_ genome (Gencode) using STAR (v2.5.3a), during which we removed duplicates and kept only uniquely mapped reads. Count files were generated by FeatureCounts (Subread v1.5.0), and BAM files were converted to bigwigs using deeptools (v3.1.3) and visualized by the UCSC genome browser. RStudio (v3.6.0) was then used to analyze counts files by DESeq2 (v1.26.0)[@loveModeratedEstimationFold2014a] to identify differentially expressed genes (DEGs) with a q-value (p-adjusted via FDR/Benjamini–Hochberg correction) less than 0.1 and a log2 fold change greater than 0.5. For all DESeq2 analyses, log2 fold changes were calculated with lfcShrink using the ashr package[@stephensFalseDiscoveryRates2016]. MA-plots were generated by ggpubr (v0.6.0), and Eulerr diagrams were generated by eulerr (v6.1.1). Boxplots and scatterplots were generated by ggpubr (v0.6.0) and ggplot2 (v3.3.2). The Upset plot was generated via the package UpSetR (v1.4.0)[@conwayUpSetRPackageVisualization2017]. Gene ontology (GO) analyses were performed by the R package enrichPlot (v1.16.2) using the biological processes setting and compareCluster.

## Chromatin immunoprecipitation followed by DNA sequecing (ChIP-seq)
ChIP-seq reads were aligned to mm10 using Bowtie1 (v1.1.2) allowing up to two mismatches. Only uniquely mapped reads were used for analysis. Peaks were called using MACS2 software (v2.1.0.20140616) using input BAM files for normalization, with filters for a q-value < 0.1 and a fold enrichment > 1. We removed “black-listed” genomic regions that often give aberrant signals. Common peak sets were obtained in R via DiffBind (v3.6.5). In the case of KDM5C ChIP-seq, _Kdm5c_-KO peaks were then subtracted from wild-type samples using bedtools (XXX). Peak proximity to genome annotations was determined by ChIPSeeker (v1.32.1). Gene ontology (GO) analyses were performed by the R package enrichPlot (v1.16.2) using the biological processes setting and compareCluster. Enriched motifs were identified using HOMER[@heinzSimpleCombinationsLineageDetermining2010]. Average binding across the genome was visualized using deeptools (v3.1.3). Bigwigs were visualized using the UCSC genome browser.

@@

## Data availability
### Published datasets
All published datasets are available at the Gene Expression Omnibus (GEO) https://www.ncbi.nlm.nih.gov/geo. Published RNA-seq datasets analyzed in this study included the male wild-type and _Kdm5c_-KO adult amygdala and hippocampus[@vallianatosMutuallySuppressiveRoles2020] (available at GEO: GSE127722) and male wild-type and _Kdm5c_-KO EpiLCs[@samantaActivationXistEvolutionarily2022] (available at GEO: GSE96797).

Previously published ChIP-seq experiments included KDM5C in wild-type and _Kdm5c_-KO EpiLCs[@samantaActivationXistEvolutionarily2022] (available at GEO: GSE96797) and mouse primary neuron cultures (PNCs) from the cortex and hippocampus[@iwaseMouseModelXlinked2016] (available at GEO: GSE61036). ChIP-seq of histone 3 lysine 4 dimethylation in male wild-type and _Kdm5c_-KO EpiLCs[@samantaActivationXistEvolutionarily2022] is also available at GEO: GSE96797. ChIP-seq of histone 3 lysine 4 trimethylation in wild-type and _Kdm5c_-KO male amygdala[@vallianatosMutuallySuppressiveRoles2020] are avilable at GEO: GSE127817.

@@

- for methods: Heatmaps of gene expression were generated using the base R functions scale and hclust and visualized using the R package ComplexHeatmap (v2.12.1).
## Discussion notes
